Associations of environmental and human factors with altered rodent populations in West Africa: A scoping review

### 2021-05-11 reports/manuscript.Rmd
draft manuscript

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

@@

Rodents represent around 40% of all mammalian species, they are important components of small mammal assemblages globally. In sub-Saharan Africa particularly they are significant contributors to agricultural crop loss in addition to being hosts for a number of zoonotic pathogens of importance to local human populations. Trapping of individual rodents, identification to species and assays for the presence of potential zoonotic pathogens provides a source of data for understanding the potential ecological and human health impact of changing small mammal populations.

@@

This was conducted as a scoping review of all rodent trapping studies identified through a systematic search of online databases for studies conducted within the United Nations West Africa subregion.

@@

4,282 records were identified from the initial search with 124 studies published between 1974 and 2021 included in narrative synthesis. The majority of studies investigated rodent ecology (56%) with the remainder investigating potential zoonotic diseases. Study methodology was comprehensively reported from 28 (23%) studies. Studies were conducted in 14 of 16 West African countries. Specific trapping habitats were not reported using standardised approaches. The included studies reported a combined 73,164 trapped small mammals, predominantly from the order Rodentia. These individuals were identified to 147 species with *Mastomys natalensis*, *Rattus rattus* and *Mastomys erythroleucus* the most commonly trapped species. 55 studies investigated potential zoonotic pathogens, 7 further studies investigated rodent pathogens. Thirty-two microorganisms were tested for including viruses, bacteria and parasites. *Mastomys natalensis* was found to be most commonly infected by *Lassa mammarenavirus* and *Bartonella sp*, *Arvicanthis niloticus* was most commonly infected with *Borrelia sp*, *Mus musculus* was most commonly infected with *Toxoplasma gondii*.

@@

This scoping review identified 124 studies, conducted in 14 countries at 1,193 study sites trapping 73,164 small rodents. Current research efforts to describe the rodent populations of West Africa has provided some information about the distribution of species across this heterogenous region and the distribution of potential pathogens within individual species. We make some recommendations for this area of research to support data sharing and re-use to better understand dynamic changes within these complex systems as a result of alterations to local habitats or climate.

@@

Beyond crop loss a further threat to local communities is the transmission of infectious diseases from rodents to humans. Zoonotic diseases (i.e. animal pathogens or microorganisms residing in animals that impact human health) are transmitted via two pathways – a direct and an indirect pathway. The direct transmission pathway involves transmission of pathogens through a rodent biting or scratching a human, or through contamination of food or water sources. The indirect pathway involves rodents acting as amplifying hosts for ectoparasites that themselves are vectors of zoonoses [@meerburg_rodent-borne_2009]. Rodent-borne pathogens vary in their scale from those with global distribution, such as the bacteria *Borrelia burgdorferi* (causing Lyme disease) and *Leptospira spp.* (casing Leptospirosis) and parasites such as *Toxoplasma gondii* (causing Toxoplasmosis) and *Leishmania spp.* (causing visceral and cutaneous Leishmaniasis). However, the majority of rodent-borne pathogens have more limited geographic distributions, viral diseases such as Hantavirus pulmonary syndrome (caused by Hantaviridae) are typically concentrated in the Americas and Lassa fever (caused by Arenaviridae) is limited to West Africa [@mills_ecologic_1998].

@@

To date there has been no comprehensive overview of rodent trapping studies conducted in West Africa. The aim of this scoping review is therefore to synthesise the knowledge on rodent distributions throughout West Africa obtained from trapping studies.

@@

  +	1. In which West African countries or regions have rodent trapping studies been performed?
  + 2. What are the stated aims of these rodent trapping studies?
  + 3. Which rodent species are being targeted for trapping activities? Are species with high potential for hosting zoonotic pathogens targeted at higher rates?
  + 4. How is the type of land use and/or land use intensity classified in the literature and are classifications comparable across studies?
  +	5. Is rodent trapping occurring in all habitat types or are some habitats targeted at higher rates?

@@

A scoping review approach was chosen to support a more descriptive and exploratory approach than using a systematic review.

  + Scoping review
  + Protocol pre-registered on the [Openscience Framework](...)

@@

  + Studies were included if:
    + Trapping of small mammals including at least one species of Rodentia.
    + Described trap used, length of trapping activity or location of trapping activity.
    + Included trapping activity from at least one West African country: Benin, Burkina-Faso, Cape Verde, Gambia, Ghana, Guinea, Guinea-Bissau, Cote d'Ivoire, Liberia, Mali, Mauritania, Niger, Nigeria, Senegal, Sierra Leone and Togo
    + Recorded as outcome genus or species of trapped individual
    + Were available in English
    + Published in a peer-reviewed journal, a pre-print manuscript or a report by a an identifiable organisation/consultancy
  + Studies were excluded if:
    + Data were duplicated from a previously included study (i.e. secondary data analysis)
    + No full text were available

@@

The following terms were searched for in OVID Medline, Web of Science (Core collection and Zoological Record), JSTOR, BioOne, African Journals Online, Global Health and the pre-print servers BioRxiv and EcoEvoRxiv:

@@

Additional articles were identified through references within included articles and reports known to the study team. Searches were completed on 2021-03-01.

@@

One reviewer screened titles, abstracts and full texts against the inclusion criteria. A random subset of each of these (10%) were reviewed by a second reviewer.

@@

Data were extracted by one reviewer. A second reviewer verified a subset (10%) of included studies. A data extraction form was maintained on a Google sheets document. Data extracted included i) study identifiers; ii) study aims; iii) trapping methodology; iv) geolocation data; v) method of speciation; vi) trapping locations and dates; vii) trapped species; viii) number of trap-nights and ix) microorganisms of interest.

@@

Where explicitly stated the study aims were extracted. For those with no recorded aims the specific aims were inferred from the introduction or conclusions of manuscript or report. The aims of the studies were categorised by the research team with studies subsequently broadly categorised as pertaining to rodent ecology research or studies on risk of zoonoses.

@@

GPS locations were extracted for the most precise location presented i.e. trap, trap-line, study-site or study region. Coordinates were extracted in the format reported and converted to decimal degrees. Where no GPS location is reported coordinates will be matched to study locations using the National Geospatial Intelligence Agency NGA GEOnet Names Server [@national_geospatial-intelligence_agency_nga_nodate] based on study site name/region and through maps presented in the manuscript.

@@

The brand, if reported, or the description of the structure of the rodent trap was obtained from each study. For studies using multiple devices all types were recorded. The method employed when setting the traps was extracted where reported. Studies that reported setting a transect line were classified as such, for studies reporting using a line of traps but not through a specified habitat the term trap line was used.

The trapping effort in trap-nights was recorded. A trap-night is the setting of a single trap for a single night. If this was reported at the habitat level or the level at which the study reported rodent captures it was classified as a complete recording of trapping effort. For studies that reported trapping effort at different levels to the capture level the recording of trap-nights was classified as incomplete. For studies reporting number of nights trapping occurred for or for where the number of traps set was stated but the number of nights not recorded they were classified as incomplete.

@@

@@

In studies investigating rodents for potential zoonoses the pathogens tested for were extracted. For studies reporting multiple potential pathogens all of the organisms were extracted. For studies solely reporting on pathogens of rodents (i.e. rodent parasites) pathogens were grouped as "Rodent pathogens". The number of rodents tested and the number of positive or negative samples were extracted alongisde the type of assay used to detect the pathogens (e.g. Polymerase Chain Reaction (PCR), Enzyme Linked ImmunoSorbent Assay (ELISA) or viral culture). Where possible pathogens were identified to the species level, however, where an assay only allows for attribution to a family of viruses or bacteria the higher order grouping was used.

@@

A total of 4,282 records were identified, with 124 studies included in narrative synthesis (see Figure \@ref(fig:prisma)). A table of the included studies and their DoIs is presented in Supplementary table 1. The earliest studies identified were from 1974 with the majority (61%) published since 2010. Most of the included studies were classified as rodent ecology research (56%) with the remainder of studies on zoonotic diseases (44%).

@@

@@

The detailed aims of the included studies are shown in Figure \@ref(fig:aims). Biodiversity of rodents was a shared aim between the two themes with 37 ecological studies and 2 zoonoses studies investigating this. The remaining aims were more disparate with ecological studies investigating rodent population dynamics, rodent species assemblages, the genetics of rodents and their behaviours. Most zoonotic disease research focussed on prevalence of pathogens in rodent populations, the risk of spillover to human populations, the identity of pathogen reservoir species with fewer studies investigating the phylogenetics of pathogen strains.

``` {r aims, fig.cap = "The aims of included studies following division into ecological and zoonotic infectious disease themes. Multiple aims were identified from each study, the number on the arrow indicates the number of distinct studies with that aim. Opening the image separately will allow you to zoom in.", echo = F, layout = "l-screen-inset shaded"}
knitr::include_graphics(here::here("figures", "study_aims.png"))
```

@@

Trapped occurred in 17 countries, however, three countries were outside of the scope of this study (Cameroon and Chad in Central Africa and Morocco in Northern Africa). At least a single trap site was recorded from 14 West African countries, with no rodent trapping reported from Gambia or Togo.

Most studies (93%) were conducted in a single country, five studies were conducted in two countries, three studies were conducted in five countries and one study was conducted across six countries. In summary 146 trapping activities were reported across the 14 West African countries. More trapping activity was reported from Senegal (19%) than any other country. Nigeria (13%), Ghana and Guinea (both 12%) were the next most frequently trapped countries, followed by Mali and Benin (both 8%), Niger and Sierra Leone (both 6%), Mauritania and the Ivory Coast (4%), Burkina Faso (3%), Liberia (2%) with Guinea-Bissau and Cape Verde recording a single trapping activity.

The trapping activity from the included studies was spread across 1,193 study sites. Thirty-one (25%) studies reported trapping only at a single study site, 45 (37%) studies trapped at between two and five study sites, the remaining 48 studies trapped at between six and 93 study sites. Trap sites were mapped to level two country administrative regions and density of sites per thousand square kilometers was calculated (see Figure \@ref(fig:density)). The areas with highest trapping densities included the capital cities of Niger, Sierra Leone, Senegal, Benin, Mali, Ghana and the Ivory Coast. Of the areas not incorporating capital cities Northern Senegal (Fatique, Thies, Saint-Louis and Kedougou), Southern Guinea (Kindia and Nzerejore), Edo and Osun States in Nigeria and Eastern Sierra Leone were the most intensively studied regions.

@@

Forty studies did not describe the structure of the trap setup within the defined study area. Twenty seven studies placed transect lines of traps through study sites with nineteen studies placing traps along a line but not clearly describing a transect across a defined site. Twenty-two studies purposefully sampled houses, with nine further studies purposefully placing traps in targeted locations (i.e. near rivers, in fields, by houses). Ten studies placed traps in a structured grid across a single or multiple specified habitat types. Three studies described a stratified or randomised approach to placing traps within a defined area. Trap setup was commonly reported in ecology studies (81%) with transect lines the most commonly used approach. Trap setup was reported in 56% of zoonoses risk studies with purposeful placement of traps in houses the most common approach (24%).

@@

Trapping effort was variably reported across included studies. Thirty-nine studies did not report any measures of trapping effort. Forty-two studies reported a measure of trapping effort at the same level of detail at which they reported rodent captures  (the unit of analysis) and were classified as complete reports. Thirty-nine studies had incomplete reports on trapping effort due to not presenting the number of trap nights at the appropriate unit of analysis (i.e. number of study nights only or total number of trap-nights). Trapping effort was more completely reported from ecology studies (51% complete information and 21% no information) compared to zoonoses risk studies (11% complete information and 44% no information).

Among studies that reported complete trap-night information the median number of trap-nights per study was 4,418 (Interquartile range (IQR) = 1,187-10,252) with a minimum of 240 trap-nights and maximum of 45,274 trap-nights. For studies reporting the number of trap-nights at the unit of analysis (typically study site or habitat type) the median number of trap-nights at each site was 360 (IQR = 246-793). In studies with incomplete reports on trap-nights there was a median of 7,200 trap-nights (IQR = 1,184-9,424) conducted, for 26 studies reporting only on the number of days or nights trapping was performed on activities typically lasted between 1-4 nights.

@@

No studies reported trap habitats with direct reference to a habitat classification scheme. Thirteen studies did not describe the local habitat where traps were placed (4 ecology, 9 zoonoses). The remaining studies used more than 170 descriptions of the habitat type in which trapping was conducted, further, the resolution of the description varied between studies with some reporting the habitat of the location of the trap and others the study site or region. The reported habitat types were combined into 30 categories (see Supplementary table 2 for the habitat dictionary) for further analysis. At least one habitat was recorded for 17,122 trap sites (95%), with two or more habitats for a single trap site recorded for 4,403 (24%) sites. Single trap sites could span multiple habitat types resulting in 22,202 habitat or trap sites. The most commonly trapped sites were in or around buildings (29%), in areas described as the rodents "natural habitat" (24%), in agricultural areas (13%) (e.g. rice fields, palm plantations), forests (8%) and in the area surrounding buildings (4%).

Habitat type was more comprehensively reported from ecology studies, 30 habitat categories were reported from ecology studies with the most commonly investigated habitats being agriculture (23%), within forests (18%) and in or around buildings (16%). Among studies on zoonoses 15 habitat categories were reported, the most commonly reported classification was described as the rodents "natural habitat" (39%), within or around buildings (38%) and in agricultural settings (7%).

@@

Thirteen included studies did not describe the method used to identify trapped rodents. Most studies (52%) relied solely on morphometric identification against taxonomic keys or museum voucher specimens. Nine studies described molecular identification of rodents with 30% using a combination of morphological and molecular techniques. In more recently published studies since 2010, 76% used molecular techniques to assist with speciation, compared with less than 15% in studies published prior to 2010.

@@

Twenty-six studies reported a numeric measure of rodent species diversity or richness, these included the Shannon Index, Simpson's Index and Chao Index. Six of these studies present a species accumulation curve to estimate the completeness of their sampling activity. A further 8 studies that did not calculate diversity indices produced species accumulation curves to measure completeness of sampling.

@@

@@

The most common pathogens assessed for with PCR were *Lassa mammarenavirus* or other Arenaviridae (31%) the bacteria *Borrelia sp.* (11.3%) and *Bartonella sp.* (6.5%) followed by Usutu virus (6.4%) and Hantaviridae (5.6%). The most common pathogens assessed for with serology were *Lassa mammarenavirus* or other Arenaviridae (78%), *Toxoplasma gondii* (10.6%), *Borrelia sp.* (6.3%) and *Leptospirosis sp.* (2.2%). The most common pathogens assessed for with histology or direct visualisation were *Borrelia sp.* (48.7%), *Schistosoma sp.* (20.4%) and other parasites. All the studies using direct culture were assessing for either *Lassa mammarenavirus* (81%) or *Leishmania sp.* (19%).

@@

Most studies tested for a single pathogen (39), with two studies testing for six pathogens, one for five pathogens, two for three pathogens and 11 for two pathogens. The most frequently tested for pathogens were *Lassa mammarenavirus* (28%) or members of the *Arenaviridae* family (16%), the spirochete bacteria *Borrelia sp.* was investigated in 8 studies, *Bartonella sp.* and *Toxoplasma gondii* were investigated in 4 each, the remaining 25 pathogens were reported in three or fewer studies.

@@

*Lassa mammarenavirus* was detected in 414 *Mastomys natalensis*, 9 *Mastomys erythroleucus*, 5 *Hylomyscus pamfi* and 3 *Mus baoulei* through PCR (demonstrating acute infection), in serological assays, evidence of prior infection or acquisition of maternal antibodies was found in 465 *Mastomys natalensis* and 361 *Mastomys sp.* individuals alongside 6 *Mus musculus* and 4 *Rattus rattus*. From direct viral culture 19 *Mastomys natalensis*, 2 *Rattus rattus* and 1 *Mus minutoides* were found to be infected with *Lassa mammarenavirus*. *Bartonella sp.* were detected in 113 *Mastomys natalensis*, 50 *Mastomys erythroleucus*, 39 *Rattus norvegicus*, 16 *Rattus rattus* and 13 individuals of 7 further species through PCR.

*Borrelia sp.* were identified via PCR in 110 individuals, 37 *Mastomys erythroleucus*, 12 *Mus musculus*, 11 *Mastomys huberti* and *Praomys daltoni*, the remaining 39 positive results were from 9 further species. In serological assays 47 *Mastomys natalensis*, 12 *Mastomys erythroleucus* and 8 *Praomys daltoni* were found to be positive, the remaining 11 positive individuals came from three further species. Most *Borrelia sp.* infections were identified through histology or direct visualisation with, 243 *Arvicanthis niloticus*, 102 *Mastomys natalensis*, and 78 *Mastomys huberti* along with a further 64 individuals from 8 species infected.

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source(here::here("scripts", "manuscript_library.r"))
```

@@

Rodents represent an estimated 40% of mammalian species and are important components of small mammal assemblages globally. In sub-Saharan Africa they are significant contributors to agricultural crop loss alongside being hosts for a number of zoonotic pathogens of importance to local human populations. Trapping of individual rodents, identification to species and assays for the presence of potential zoonotic pathogens produces important data for understanding the potential ecological and human health impact of changing small mammal populations.

@@

A scoping review of all rodent trapping studies identified through a systematic search of online databases for studies conducted within the United Nations West Africa sub-region was performed.

@@

4,282 records were identified from the initial search with 124 studies published between 1974 and 2021 included in narrative synthesis. Studies were conducted in 14 West African countries. The majority of studies investigated rodent ecology (56%) with the remainder investigating potential zoonotic diseases. Study methodology was comprehensively reported from 28 (23%) studies. Specific trapping habitats were not reported using standardised approaches. The included studies reported 73,164 trapped small mammals, predominantly from the order Rodentia. These individuals were identified to 147 species with *Mastomys natalensis*, *Rattus rattus* and *Mastomys erythroleucus* the most common. 55 studies investigated potential zoonotic pathogens, 7 further studies investigated rodent pathogens. Thirty-two microorganisms were tested for including viruses, bacteria and parasites. *Mastomys natalensis* was found to be most commonly infected by *Lassa mammarenavirus* and *Bartonella sp*, *Arvicanthis niloticus* was most commonly infected with *Borrelia sp* and *Mus musculus* was most commonly infected with *Toxoplasma gondii*.

@@

This scoping review identified 124 studies, conducted in 14 countries at 1,193 study sites trapping 73,164 small rodents. Current research to describe the rodent populations of West Africa have provided some information about the distribution of species across this heterogenous region and the distribution of potential pathogens within individual species. We make recommendations for subsequent research to enable wider data sharing and re-use to to support meta-analysis of obtained data to address questions on rodent ecology and zoonotic disease risk.

@@

Alongside crop loss a further risk to local communities is the transmission of infectious diseases from rodents to humans. Zoonotic diseases (i.e. animal pathogens or microorganisms residing in animals that impact human health) are transmitted via two pathways – a direct and an indirect pathway. The direct transmission pathway involves transmission of pathogens through a rodent biting or scratching a human, or through contamination of food or water sources. The indirect pathway involves rodents acting as amplifying hosts for ectoparasites that themselves are vectors of zoonoses [@meerburg_rodent-borne_2009]. Rodent-borne pathogens vary in their scale from those with global distribution, such as the bacteria *Borrelia burgdorferi* (causing Lyme disease) and *Leptospira spp.* (casing Leptospirosis) and parasites such as *Toxoplasma gondii* (causing Toxoplasmosis) and *Leishmania spp.* (causing visceral and cutaneous Leishmaniasis). However, the majority of rodent-borne pathogens have more limited geographic distributions, viral diseases such as Hantavirus pulmonary syndrome (caused by Hantaviridae) are typically concentrated in the Americas and Lassa fever (caused by Arenaviridae) is limited to West Africa [@mills_ecologic_1998].

@@

To date there has been no comprehensive overview of rodent trapping studies conducted in West Africa. The aim of this scoping review is therefore to systematically map the research conducted in this area, to identify currently used methods, gaps in current knowledge and to synthesise the knowledge on rodent distributions throughout West Africa obtained from included trapping studies.

@@

  1. In which West African countries or regions have rodent trapping studies been performed?
  2. What are the stated aims of these rodent trapping studies?
  3. Which rodent species are being targeted for trapping activities? Are species with high potential for hosting zoonotic pathogens targeted at higher rates?
  4. How is the type of land use and/or land use intensity classified in the literature and are classifications comparable across studies?
  5. Is rodent trapping occurring in all habitat types or are some habitats targeted at higher rates?

@@

A scoping review approach was used to identify and map the available information on rodent assemblages, abundance and diversity across West Africa. This approach is recommended when examining how research is conducted within a topic area [@munn_systematic_2018]. The review followed the recommended appreach from the PRISMA extension for Scoping Review guidance [@tricco_prisma_2018]. The protocol for this review was not pre-registered.

@@

  + Studies were included if:
    + Trapping of small mammals including at least one species of Rodentia.
    + Described trap used, length of trapping activity or location of trapping activity.
    + Included trapping activity from at least one West African country: Benin, Burkina-Faso, Cape Verde, Gambia, Ghana, Guinea, Guinea-Bissau, Cote d'Ivoire, Liberia, Mali, Mauritania, Niger, Nigeria, Senegal, Sierra Leone and Togo
    + Recorded as outcome genus or species of trapped individual
    + Published in a peer-reviewed journal, a pre-print manuscript or a report by a an identifiable organisation/consultancy
  + Studies were excluded if:
    + Data were duplicated from a previously included study (i.e. secondary data analysis)
    + No full text were available
    + Not available in English

@@

The following terms as keywords were searched for in OVID Medline, Web of Science (Core collection and Zoological Record), JSTOR, BioOne, African Journals Online, Global Health and the pre-print servers BioRxiv and EcoEvoRxiv:

@@

Additional articles were identified through references within included articles and reports known to the study team. No time constraints were placed on searches. Searches were completed on 2021-03-01. The exploded search terms and links to the respective web portals are provided in supplementary material.

@@

One reviewer screened titles, abstracts and full texts against the inclusion and exclusion criteria. A random subset of each of these (10%) were reviewed by a second reviewer.

@@

Data from eligible studies were extracted using a standardised data abstraction tool designed for this study on a subset of eligible studies and subsequently refined. Data were extracted by a single reviewer. A second reviewer verified a random subset (10%) of included studies. A data extraction form was maintained on a Google sheets document. Data extracted included i) study identifiers; ii) study aims; iii) trapping methodology; iv) geolocation data; v) method of speciation; vi) trapping locations and dates; vii) trapped species; viii) number of trap-nights and ix) microorganisms of interest. The data extraction forms and data dictionaries used during the analysis are reproduced in the supplementary material.

@@

When explicitly stated the study aims were extracted, for those with no specified aims the aims were inferred from the introduction or conclusions of the manuscript or report. The aims of studies were categorised with a subsequent higher order allocation to rodent ecology research or studies on risk of zoonoses.

@@

GPS locations were extracted for the most precise location presented i.e. trap, trap-line, study-site or study region. Coordinates were extracted in the format reported and converted to decimal degrees. Where no GPS location was reported coordinates were matched to study locations using the National Geospatial Intelligence Agency NGA GEOnet Names Server [@national_geospatial-intelligence_agency_nga_nodate] based on study site name/region and comparison to maps presented in the manuscript (when available).

@@

The brand, if reported, or the description of the structure of the rodent trap was obtained from each study. For studies using multiple devices all types were recorded. The method employed when setting the traps was also extracted, studies that reported setting a transect line were classified as such, for studies reporting using a line of traps but not through a specified habitat the term trap line was used.

The trapping effort in trap-nights (the setting of a single trap for a single night) was recorded. If this was reported at the habitat level or the level at which the study reported rodent captures it was classified as a complete recording of trapping effort. For studies that reported trapping effort at different levels to the capture level the recording of trap-nights was classified as incomplete. For studies reporting number of nights trapping occurred for or for where the number of traps set was stated but the number of nights not recorded they were classified as incomplete.

@@

#### Genus/species' trapped

For studies reporting on all trapped species (i.e. not those only reporting on specific species of interest) that do not report absence at study sites all species trapped within the study will be recorded at all study sites with explicitly zero observations.

@@

In studies investigating rodents for potential zoonoses all pathogens tested for were extracted. The number of rodents tested and the number of positive or negative samples were extracted alongside the type of assay used to detect the pathogens (e.g. Polymerase Chain Reaction (PCR), Enzyme Linked ImmunoSorbent Assay (ELISA) or viral culture). Where possible pathogens were identified to the species level, however, where an assay only allows for attribution to a family of viruses or bacteria the higher order grouping was used (i.e. PCR using a non-specific arenavirus primer).

#### Risk of bias

No formal risk of bias assessment tool was used.

@@

A total of 4,282 records were identified, with 124 studies included in narrative synthesis (see Figure \@ref(fig:prisma)). A summary table of the included studies and the citations of the study or report are presented in supplementary table 1. The earliest studies identified were from 1974 with the majority (61%) published since 2010. Most of the included studies were classified as rodent ecology research (56%) with the remainder of studies on zoonotic diseases (44%).

@@

Table 1 summarises the included studies by publication year, and methodology with grouping by study aim.

``` {r table-1, echo=F}
studies <- read_rds(here("data_clean", "studies.rds"))

studies %>%
  mutate(decade_publication = case_when(year_publication < 1980 ~ "1970-79",
                                        year_publication < 1990 ~ "1980-89",
                                        year_publication < 2000 ~ "1990-99",
                                        year_publication < 2010 ~ "2000-09",
                                        year_publication < 2020 ~ "2010-19",
                                        year_publication < 2030 ~ "2020-",
                                        TRUE ~ "other"),
         geolocation_level = case_when(geolocation_level == "island" ~ "Study site",
                                       geolocation_level == "none" ~ "No geolocation",
                                       geolocation_level == "study" ~ "Study",
                                       geolocation_level == "study_region" ~ "Study",
                                       geolocation_level == "study_site" ~ "Study site",
                                       geolocation_level == "trap_line" ~ "Specific trap site",
                                       geolocation_level == "trap_site" ~ "Specific trap site",
                                       geolocation_level == "trap_site, study_site" ~ "Specific trap site",
                                       TRUE ~ "other"),
         speciation = case_when(speciation == "molecular" ~ "Molecular",
                                speciation == "morphological" ~ "Morphological",
                                speciation == "morphological, molecular" ~ "Morphological and molecular",
                                speciation == "not_stated" ~ "Not stated",
                                TRUE ~ "other"),
         pathogen = case_when(pathogen == "Yes, second paper" ~ "Yes",
                               TRUE ~ pathogen),
         country = sub('(^[^_]+_[^_]+)_(.*)$', '\\2', unique_id),
         country = case_when(country == "multiple" ~ "Multiple",
                             TRUE ~ countryname(sub("_2", "", country), destination = "un.name.en"))) %>%
  dplyr::select(country, decade_publication, aim, repeated_visit, geolocation_level,
                speciation, species_accumulation, diversity_measurement, trapping_effort, pathogen) %>%
  tbl_summary(by = aim,
              label = c(country ~ "Country",
                        decade_publication ~ "Publication year",
                        repeated_visit ~ "Repeat study visits - Yes",
                        geolocation_level ~ "Level of geolocation",
                        speciation ~ "Method of speciation",
                        species_accumulation ~ "Use of a species accumulation curve - Yes",
                        diversity_measurement ~ "Use of a measure of species diversity - Yes",
                        trapping_effort ~ "Reporting of trapping effort",
                        pathogen ~ "Reporting of potential pathogens")) %>%
  add_overall() %>%
  as_gt() %>%
  tab_header(title = "Table 1: Characteristics of included studies summarised by study aim") %>%
  tab_options(table.font.size = px(12),
              data_row.padding = px(2),
              container.height = 600) %>%
  fmt_missing(columns = everything(), missing_text = " ")
```

@@

The detailed aims of the included studies are shown in Figure \@ref(fig:aims). Biodiversity of rodents was a shared aim between the two themes with 37 ecological studies and 2 zoonoses studies investigating this.

``` {r aims, fig.cap = "The aims of included studies following division into ecological and zoonotic infectious disease themes. Multiple aims were identified from each study, the number on the arrow indicates the number of distinct studies with that aim.", echo = F, layout = "l-screen-inset shaded"}
knitr::include_graphics(here::here("figures", "study_aims.png"))
```

@@

Rodent trapping occurred in 17 countries, three countries were outside of the scope of this study (Cameroon and Chad in Central Africa and Morocco in Northern Africa). At least a single trap site was recorded from 14 West African countries. No rodent trapping was identified from Gambia or Togo.

Most studies (92%) were conducted in a single country, six studies were conducted in two countries, three studies were conducted in five countries and one study was conducted across six countries. In summary 146 trapping activities were reported across the 14 West African countries.

The trapping activity from the included studies was performed at 1,193 study sites. Thirty-one (25%) studies reported trapping only at a single study site, 45 (37%) studies trapped at between two and five study sites, the remaining 48 studies trapped at between six and 93 study sites. Trap sites were mapped to level two country administrative regions and density of sites per thousand square kilometers was calculated (see Figure \@ref(fig:density)). The areas with highest trapping densities included the capital cities of Niger (Niamey), Sierra Leone (Freetown), Senegal (Dakar), Mali (Bamako) and Ghana (Accra) and the largest cities of Ivory Coast (Abidjan) and Benin (Cotonou). Of the areas not incorporating these cities; Northern Senegal (Fatique, Thies, Saint-Louis and Kedougou), Southern Guinea (Kindia and Nzerejore), Edo and Osun States in Nigeria and Eastern Sierra Leone were the most intensively studied regions.

@@

Forty studies did not describe the trap setup within a defined study area. Twenty seven studies placed transect lines of traps through study sites with nineteen studies placing traps along a line but not clearly describing a transect across a defined habitat. Twenty-two studies purposefully sampled domestic structures (i.e. homes), with nine further studies purposefully placing traps in targeted locations (i.e. near rivers, in fields, by houses). Ten studies placed traps in a structured grid across a single or multiple specified habitat types. Three studies described a stratified or randomised approach to placing traps within a defined area. Trap setup was commonly reported in ecology studies (81%) with transect lines the most commonly used approach. Trap setup was reported in 56% of zoonoses risk studies with purposeful placement of traps in houses the most common method (24%).

@@

Thirty-nine studies did not report any measures of trapping effort. Forty-two studies reported a measure of trapping effort at the same level of detail at which they reported rodent captures  (the unit of analysis) and were classed as complete reports. Thirty-nine studies were classed as incomplete reports on trapping effort due to not presenting the number of trap nights at the same scale as the unit of analysis (i.e. number of study nights only or total number of trap-nights). Trapping effort was more completely reported from ecology studies (51% complete information and 21% no information) compared to zoonoses risk studies (11% complete information and 44% no information).

Among studies that reported complete trap-night information the median number of trap-nights per study was 4,418 (Interquartile range (IQR) = 1,187-10,252) with a minimum of 240 trap-nights and maximum of 45,274 trap-nights. For studies reporting the number of trap-nights at the unit of analysis (typically study site or habitat type) the median number of trap-nights at each site was 360 (IQR = 246-793). In studies with incomplete reports on trap-nights there was a median of 7,200 trap-nights (IQR = 1,184-9,424) conducted. For 26 studies reporting only on the number of days or nights trapping was performed the trapping activity typically lasted between 1-4 nights.

@@

No studies reported trap habitats with direct reference to a habitat classification scheme. Thirteen studies did not describe the local habitat in which traps were placed (4 ecology, 9 zoonoses). The remaining studies used more than 170 descriptions of the habitat type in which trapping was conducted, further, the scale of habitat of the description varied. The extracted habitat types were grouped into 30 categories (see Supplementary table 2 for the habitat dictionary) for further analysis. At least one habitat was recorded for 17,122 trap sites (95%), with two or more habitats for a single trap site recorded for 4,403 (24%) sites. Single trap sites could span multiple habitat types resulting in 22,202 habitat or trap sites. The most commonly trapped sites were in or around buildings (29%), in areas described as the rodents "natural habitat" (24%), in agricultural areas (13%) (e.g. rice fields, palm plantations), forests (8%) and in the area surrounding buildings (4%).

Habitat type was more comprehensively reported from ecology studie and were most commonly agricultural (23%), within forests (18%) and in or around buildings (16%). The most commonly reported habitats from zoonoses studies were the rodents "natural habitat" (39%), within or around buildings (38%) and in agricultural settings (7%).

@@

Thirteen included studies did not describe the method used to identify trapped rodents. Most studies (52%) relied solely on morphometric identification against taxonomic keys or museum voucher specimens. Nine studies described molecular identification of rodents with 30% using a combination of morphological and molecular techniques. In more recently published studies since 2010, 76% used molecular techniques to aid speciation, compared with less than 15% in studies published prior to 2010.

@@

Twenty-six studies reported a numeric measure of rodent species diversity or richness, the indices used included the Shannon Index, Simpson's Index and Chao Index. Six of these studies present a species accumulation curve to estimate the completeness of their sampling activity. A further 8 studies that did not present diversity indices produced species accumulation curves to measure completeness of sampling.

@@

``` {r pathogen-table, echo=F}
readRDS(here::here("tables", "pathogen_table.rds")) %>%
  ungroup() %>%
  gt::gt(rowname_col = "Pathogen tested") %>%
  gt::tab_header(title = "Table 2: The number of studies testing for each microorganism or group of microorganisms by assay type") %>%
  tab_options(table.font.size = px(12),
              data_row.padding = px(2)) %>%
  gt::fmt_missing(columns = everything(), missing_text = " ")
```

@@

The most common pathogens assessed for with PCR were *Lassa mammarenavirus* or other Arenaviridae (31%) the bacteria *Borrelia sp.* (11.3%) and *Bartonella sp.* (6.5%) followed by Usutu virus (6.4%) and Hantaviridae (5.6%). The most common pathogens assessed for with serology were *Lassa mammarenavirus* or other Arenaviridae (78%), *Toxoplasma gondii* (10.6%), *Borrelia sp.* (6.3%) and *Leptospirosis sp.* (2.2%). The most common pathogens assessed for with histology or direct visualisation were *Borrelia sp.* (48.7%), *Schistosoma sp.* (20.4%) and other parasites. All  studies using direct culture were investigating either *Lassa mammarenavirus* (81%) or *Leishmania sp.* (19%).

@@

Most studies tested for a single pathogen (39), with 16 studies testing for two or more pathogens. The most frequently tested for pathogens were *Lassa mammarenavirus* (28%) or members of the *Arenaviridae* family (16%), the spirochete bacteria *Borrelia sp.* was investigated in 8 studies, *Bartonella sp.* and *Toxoplasma gondii* were investigated in 4 each, the remaining 25 pathogens were reported in three or fewer studies.

@@

*Lassa mammarenavirus* was detected in 414 *Mastomys natalensis*, 9 *Mastomys erythroleucus*, 5 *Hylomyscus pamfi* and 3 *Mus baoulei* through PCR (demonstrating acute infection). In serological assays, evidence of prior infection or acquisition of maternal antibodies to *Lassa mammarenavirus* was found in; 465 *Mastomys natalensis*, 361 *Mastomys sp.* and 6 *Mus musculus* and 4 *Rattus rattus*. From direct viral culture 19 *Mastomys natalensis*, 2 *Rattus rattus* and 1 *Mus minutoides* were found to be infected with *Lassa mammarenavirus*. *Bartonella sp.* were detected in; 113 *Mastomys natalensis*, 50 *Mastomys erythroleucus*, 39 *Rattus norvegicus*, 16 *Rattus rattus* and 13 individuals of 7 further species through PCR.

*Borrelia sp.* infections were identified through PCR in 110 individuals; 37 *Mastomys erythroleucus*, 12 *Mus musculus*, 11 *Mastomys huberti* and *Praomys daltoni*, the remaining 39 positive results were from 9 further species. In serological assays; 47 *Mastomys natalensis*, 12 *Mastomys erythroleucus* and 8 *Praomys daltoni* were found to be positive, the remaining 11 positive individuals came from three further species. Most *Borrelia sp.* infections were identified through histology or direct visualisation with, 243 *Arvicanthis niloticus*, 102 *Mastomys natalensis*, and 78 *Mastomys huberti* along with a further 64 individuals from 8 species infected.
